### 2017-06-28 motivation.tex
Example formatting.
Motivation Edited.

In this section, we motivate our ideas via an example written in
\txnimp - a functional language equipped with a \C{DB} monad to
compose database computations. Besides the usual \C{bind} and
\C{return} combinators, monad offers a \C{txn\_do} combinator that
executes a database computation and returns the result.

\begin{figure}
\centering
\begin{tabular}{l||l}
\begin{ocaml}
let new_order_txn d_id c_d_id c_id
      (ireqs: item_req list) = atomically do
  dist <- SQL.select1 District
               (fun d -> d.d_id = d_id);
  let o_id = dist.d_next_o_id;
  SQL.update District 
    (fun d -> 
      {d with d_next_o_id =d_next_o_id + 1})
    (fun d -> d.d_id = d_id );
  SQL.insert Order {o_id=o_id;  o_d_id=d_id; 
                    o_c_id=c_id; o_ol_cnt=S.size ireqs; };
  SQL.insert New_order {no_o_id=o_id; no_d_id=d_id};
  foreach ireqs @@ fun ireq -> do
    stk <- SQL.select1 Stock
             (fun s -> s.s_i_id = ireq.ol_i_id);
    let s_qty' = if stk.s_qty >= ireq.ol_qty + 10 
                then stk.s_qty - ireq.ol_qty
                else stk.s_qty - ireq.ol_qty + 91;
    SQL.update Stock
       (fun s -> {s with s_qty = s_qty'})
       (fun s -> s.s_i_id = ireq.ol_i_id);
    SQL.insert Order_line {ol_o_id=o_id; ol_d_id=d_id; 
        ol_i_id=ireq.ol_i_id; ol_qty=ireq.ol_qty}

\end{ocaml}
&
\begin{ocaml}
  txn$\langle$'Wd2'$\rangle${
    if (B $\ge$ a2) {
      B := B - a2
    }
  }
\end{ocaml}
\\
\end{tabular}
\caption{\small Concurrent withdraw transactions}
\label{fig:motiv-eg-1}
\vspace*{-10pt}
\end{figure}

Consider an implementation of a banking application that admits
concurrent withdraw transactions on an account balance (\C{B}), as
shown in Fig.~\ref{fig:motiv-eg-1}. If the initial balance (\C{k}) in
the account is enough to perform both withdraws, then the final
balance, after both transactions commit, is expected to reflect the
effects of both withdraws. The pre- and post-conditions in
Fig.~\ref{fig:motiv-eg-1} reflect this expectation. Indeed, invariants
are guaranteed to hold if both withdraw transactions are serialized,
making \iso{Serializable} isolation ({\sc ser}) level a sufficient
condition to preserve invariants. But, is {\sc ser} necessary?

>>>>

In this section, we motivate our ideas via an example written in
\txnimp - a functional language equipped with a \C{DB} monad to
compose database computations. Besides the usual \C{bind} and
\C{return} combinators, monad offers an \C{atomically} combinator that
executes a database computation and returns the result.

\begin{figure}
\centering
\begin{ocaml}
let new_order_ d_id c_id item_reqs = atomically do
  dist <- SQL.select1 District (fun d -> d.d_id = d_id);
  let o_id = dist.d_next_o_id;
  SQL.update District (fun d -> {d with d_next_o_id =d_next_o_id + 1})
                      (fun d -> d.d_id = d_id );
  SQL.insert Order {o_id=o_id;  o_d_id=d_id; 
                    o_c_id=c_id; o_ol_cnt=S.size item_reqs; };
  foreach item_reqs @@ fun item_req -> do
    stk <- SQL.select1 Stock (fun s -> s.s_i_id = item_req.ol_i_id);
    let s_qty' = if stk.s_qty >= item_req.ol_qty + 10 
                then stk.s_qty - item_req.ol_qty 
                else stk.s_qty - item_req.ol_qty + 91;
    SQL.update Stock (fun s -> {s with s_qty = s_qty'}) 
                     (fun s -> s.s_i_id = item_req.ol_i_id);
    SQL.insert Order_line {ol_o_id=o_id; ol_d_id=d_id; 
                           ol_i_id=item_req.ol_i_id; ol_qty=item_req.ol_qty}

\end{ocaml}
\caption{\small Concurrent withdraw transactions}
\label{fig:new_order_code}
\vspace*{-10pt}
\end{figure}

Fig.~\ref{fig:new_order_code} shows a simplified version of the TPC-C
\C{new\_order} transaction written in \txnimp (the \C{do} syntax ala
Haskell is used for convenience). TPC-C is an Online Transaction
Processing (OLTP) benchmark that models an order-processing system for
a wholesale parts supply business. The business logic is captured in 5
database transactions that operate on 9 tables. \C{new\_order} is one
such trasaction that operates on \C{District}, \C{Order},
{New\_order}, \C{Stock}, and \C{Order\_line} tables. The transaction
acts on the behalf of a customer, whose id is \C{c\_id}, to place a
new order for the given set of items (\C{item\_reqs}), to be served by
a warehouse under the district identified by \C{d\_id}. The
transaction does so by invoking appropriate SQL functionality,
captured in Fig.~\ref{fig:new_order_code} by various calls to
functions under the \C{SQL} module. All \C{SQL} functions take the
table name (a nullary constructor) as their first argument. The
higher-order \C{SQL.select1} function accepts a boolean function that
describes the selection criteria, and returns any record that meets
the criteria (it models the SQL query \C{SELECT \ldots\xspace LIMIT
1}). \C{SQL.update} also accepts a boolean function (3rd argument) to
select the records to be updated. Its 2nd argument is a function that
maps each selected record to a new (updated) record. \C{SQL.insert}
inserts a given record into the specified table in the database.

The \C{new\_order} transaction inserts a new \C{Order} record, whose
id is the sequence number of the next order under the given district
(\C{d\_id}). The sequence number is stored in the corresponding
\C{District} record, and updated each time a new order added to the
system. Since each order may request multiple items (\C{item\_reqs}),
an \C{Order\_line} record is created for each requested item to relate
the order with the item. Each item has a corresponding record in the
\C{Stock} table, which keeps track of the quantity of the item left in
stock (\C{s\_qty}). The quantity is updated by the transaction to
reflect the processing of new order (if the stock quantity falls below
10, it is automatically replenished by 91).

TPC-C defines multiple invariants, called \emph{consistency
conditions}, over the state of the application in the database. One
such consistency condition is the requirement that for a given order
\C{o}, the order-line-count field (\C{o.o\_ol\_cnt}) should reflect
the number of order lines under the order, i.e., the number of
\C{Order\_line} records whose \C{ol\_o\_id} is equal to \C{o.o\_id}.
In a sequential exectution, it is more-or-less easy to see how this
condition condition is preserved by the \C{new\_order} transaction by
examining Fig.~\ref{fig:new_order_code}. A new \C{Order} record is
added with its \C{o\_id} distinct from the existing order ids, and its
\C{o\_ol\_cnt} set equal to the size of \C{item\_reqs} set. The
\C{foreach} loop runs once for each \C{item\_req}, adding a new
\C{Order\_line} record for each requested item, with its \C{ol\_o\_id}
field set to \C{o\_id}. Thus, by the end of the loop, the number of
\C{Order\_line} records in the database, whose \C{ol\_o\_id} is equal
to \C{o\_id}, is equal to the size of the \C{item\_req} set, which
inturn is equal to the \C{Order} record's \C{o\_ol\_cnt} field. Thus
the consistency condition is preserved.

While the aforementioned reasoning is simple to perform manually, the
sequential verification of database programs, such as the one in
Fig.~\ref{fig:new_order_code}, is difficult to automate owing to the
presence of loops and collections. Thus verifying serializable
transactions is at least as difficult as reasoning about sequential
imperative programs. Nonetheless, program logics and automation
techniques developed to reason about imperative programs can be used
to reason about sequential transactions, perhaps with reasonable
success.

concurrent withdraw transactions on an account balance (\C{B}), as
shown in Fig.~\ref{fig:motiv-eg-1}. If the initial balance (\C{k}) in
the account is enough to perform both withdraws, then the final
balance, after both transactions commit, is expected to reflect the
effects of both withdraws. The pre- and post-conditions in
Fig.~\ref{fig:motiv-eg-1} reflect this expectation. Indeed, invariants
are guaranteed to hold if both withdraw transactions are serialized,
making \iso{Serializable} isolation ({\sc ser}) level a sufficient
condition to preserve invariants. But, is {\sc ser} necessary?
